Index measures for oak decline severity using phenotypic descriptors

### 2019-12-15 manuscript/manuscript.Rmd
added a couple of disscussion points. Spell check

**Keywords:** acute oak decline, chronic oak decline, phenotyping, machine learning, unsupervised learning, random forest

@@

Descriptors were collected relating to the trees shape, size, crown condition and the presence of biotic agents including insects, bacteria and fungi.
All trees were manually assigned a decline status category; control, AOD, COD or remission prior to collection of phenotypic descriptors.
The remission group included trees that had shown stem bleeds associatied with AOD in the past but were not showing signs of stem bleeds at the time of assessment.

@@

Other categorical descriptors included social class, diaback type, main crown dieback location, stem epicormic growth, branch epicormic growth, extent of insect damage, insect damage type, extent of crown mildew and the presence or absence of pruning damage or branch loss from the main stem.
See supplementary table # for category details.

@@

All cardinal assessments were averaged to give a single value per tree.
The crown density ($CD$) values erranously collected at Attingham and Langdale were converted to crown transparency ($t = 100 - CD$)
All descriptors collected in cm were converted to mm.
The tap tests that were positive for hollow stem tissue were converted to present and absent renamed dead stem tissue.
The canopy closure descriptor was converted from yes and now to present and absent.
Finally, the stem symptoms sizes for active bleeds, black staining and callused wounds were averaged and a frequency per tree collected.
After preparation there were a total of 31 raw descriptors for analysis.

@@

The interactions of key raw descriptors identified from the overall importance scores and their associations with the decline indexes was investigated by using simulated descriptor values to construct index model responses surfaces for specific cases.
This approach was similar to response surface methodology without the need for probabalistic approximation due to the low computation time required for index calculation (ref).

For the PDI, missing crown and crown transparency were compared for example cases of a healthy, a moderate and a severly tree.
The example cases included the trees with the minimum and maximum PDI values and the tree with the PDI value closest to the median for the healthy, severe and moderately declining trees respectively.
The value ranges for the missing crown (`r glue("{min(site_corrected_analysis_suitable_data[['Missing crown (%)']])}-{max(site_corrected_analysis_suitable_data[['Missing crown (%)']])}%")`) and crown transparency (`r glue("{min(site_corrected_analysis_suitable_data[['Crown transparency (%)']])}-{max(site_corrected_analysis_suitable_data[['Crown transparency (%)']])}%")`) descriptors divided into 100 increments.
For a given case, each combination of theses simulated descriptor values were used to calculate a PDI value from the supervised random forest regression model, building a model response surface.

@@

The same approach was used for the DAI using also a comparison between active bleed size and black staining size as well as missing crown and crown transparency.
The upper boundary for the ranges of active bleed size and black stain size was taken as the center value between the maximum and minimum and reduced by a factor of eight to give ranges of `r glue("{symptom_ranges$minimum[1]}-{signif(symptom_ranges$center[1] / 8,digits = 3)}mm")` and `r glue("{symptom_ranges$minimum[2]}-{signif(symptom_ranges$center[2] / 8,digits = 3)}mm")` respectively and enable more effective visualisation.
Cases included neutral, moderate and severe with PDI scores greater than 0.5 and DAI scores closest to 0, 0.5 and 1 respectively for AOD and 0, -0.5 and -1 respectively for COD.

@@

A scatter plot of the calculated PDI and DAI scores is shown in Figure # for all 174 trees for which phenotypic descriptor data was collected.
It can clearly be seen that the control trees group towards the lower end of the PDI scale and the sympotmatic COD and AOD groups towards the higher end.
This gave a significant difference when PDI of the control and symptomatic (AOD and COD) decline statuses were compared using an independent two sample t-test (`r glue("t({PDI_status_ttest$parameter}) = {signif(PDI_status_ttest$statistic,digits = 3) %>% abs()}, p < 0.001")`) with group means of `r signif(PDI_status_ttest$estimate1,digits = 3)` and `r signif(PDI_status_ttest$estimate2,digits = 3)` for non-symptomatic and symptomatic statuses respectively.

@@

LIME identified crown volume, condition transparency and radius as the top four descriptors contributing to the PDI in each of the healthy, moderate and severe decline cases using lime (Table #).
These descriptors were also the same top descriptors in the overall model importance lists for the PDI.
Increasing decline severity was associated with a reduction in the size and condition of the crown.
For instance, declining trees were found to have an estimated crown volume below 841m^3^.
Healthy trees were found to have a crown condition greater than 71.3%, moderately declining trees between 47.6% and 71.3% and severly declining trees a condition below 23.9%.

@@

The DAI is a much more complex index with more descriptors contributing to it than the PDI such as bleed prevalence, diameter at breast height and crown condition descriptors. 
Simulated response surfaces for the DAI are shown in Figure c. 
These response surfaces show circumstances under which bleed prevalence related descriptors (Figures #a, b & c) and crown condition descriptors (Figures dd, e & f) change for non-symptomatic AOD or COD trees (Figures #a & d respectively), moderate AOD or COD (Figures #b & e respectively) and severe AOD or COD (Figures #c & f respectively). 
Increasing active bleed, staining size and frequency contributed to greater positive values of the DAI and more severe AOD. 
Also higher DAI values could be obtained from trees with greater stature (Figure #c). 
Crown transparency and missing crown have no influence on the DAI for trees of neutral decline type (Figure #d).
Similarly to the PDI, trees with smaller stature and poorer crown condition give lower DAI values (Figures #e & f) and therefore more severe COD.
The relationship of relative tree stature and the acuteness of the decline likely reflects the period of time under which these trees have been afflict. 
AOD occurs over a much shorter period compared COD and therefore has less time in which to inhibit growth of the tree as a whole.

@@

### Suitablity of the decline indexes for describing decline severity and type

- Suitability of collected phenotypic descriptors and decline indexes for describing decline status
- Index thresholds for category definitions (healthy, declining, severe decline etc.)
- Validation of decline scores - chicken and paradox
- Refined definition of remission - repeat measurement

This plot highlights the inaccuracies in manual field classifications of decline status with several trees classified as healthy controls that, when compared quantitatively to other trees, are in fact
in a decline state. 
Most notably this has occurred in a number of trees at Attingham, Hatchlands and Winding Wood. 
Also notable from Figure 3 is the disparity between the relative health of the healthier trees across the sites. 
The two dimensional space provided by these phenotypic indexes, for measuring decline severity is shown in Figure 1. 
The healthier trees at sites such as Attingham and Hatchlands were found to be in a more declined state than those at sites such as Langdale and Chestnuts.
The phenotypic descriptors used to calculate these indexes are easily measurable in the field and require no specialist equipment. 
They allow sensitive comparisons to be made for the same tree between years as well as comparisons of trees between different locations.

@@

- reasons for selecting machine learing approach
- use of unsupervised over supervised analysis
- Semi-supervised can be used if unsupervised model is unable to pull out decline severity scale
- For given context, descriptors not contributing to the indexes could be removed
- Concession of model interpretability to predictive capability by using a machine learning approach
- seen as black box but LIME and response surfaces enable interpretation and therefore trust
- LIME can be applied to out of place cases
- Important descriptors identified model-wide aren't necessarily relevant to localised cases.

A machine learning approach was used to derive these measures as it allows the complex interactions and trends between phenotypic descriptors to be modelled from which the decline indexes are calculated. 
This would likely be oversimplified using conventional modelling strategies [@breiman2001b].

### Applications and wider scaleablity

@@

which could have great utility in long term forest health monitoring for both forestry management and research.

>>>>

**Keywords:** acute oak decline, chronic oak decline, phenotyping, machine learning, unsupervised learning, random forest, decline spectrum

@@

Descriptors were collected relating to the trees shape, size, crown condition and the presence of biotic agents including insects, bacteria and fungi.
All trees were manually assigned a decline status category; control, AOD, COD or remission prior to collection of phenotypic descriptors.
The remission group included trees that had shown stem bleeds associated with AOD in the past but were not showing signs of stem bleeds at the time of assessment.

@@

Other categorical descriptors included social class, dieback type, main crown dieback location, stem epicormic growth, branch epicormic growth, extent of insect damage, insect damage type, extent of crown mildew and the presence or absence of pruning damage or branch loss from the main stem.
See supplementary table # for category details.

@@

All cardinal assessments were averaged to give a single value per tree.
The crown density ($CD$) values erroneously collected at Attingham and Langdale were converted to crown transparency ($t = 100 - CD$)
All descriptors collected in cm were converted to mm.
The tap tests that were positive for hollow stem tissue were converted to present and absent renamed dead stem tissue.
The canopy closure descriptor was converted from yes and now to present and absent.
Finally, the stem symptoms sizes for active bleeds, black staining and callused wounds were averaged and a frequency per tree collected.
After preparation there were a total of 31 raw descriptors for analysis.

@@

The interactions of key raw descriptors identified from the overall importance scores and their associations with the decline indexes was investigated by using simulated descriptor values to construct index model responses surfaces for specific cases.
This approach was similar to response surface methodology without the need for probabilistic approximation due to the low computation time required for index calculation (ref).

For the PDI, missing crown and crown transparency were compared for example cases of a healthy, a moderate and a severely tree.
The example cases included the trees with the minimum and maximum PDI values and the tree with the PDI value closest to the median for the healthy, severe and moderately declining trees respectively.
The value ranges for the missing crown (`r glue("{min(site_corrected_analysis_suitable_data[['Missing crown (%)']])}-{max(site_corrected_analysis_suitable_data[['Missing crown (%)']])}%")`) and crown transparency (`r glue("{min(site_corrected_analysis_suitable_data[['Crown transparency (%)']])}-{max(site_corrected_analysis_suitable_data[['Crown transparency (%)']])}%")`) descriptors divided into 100 increments.
For a given case, each combination of theses simulated descriptor values were used to calculate a PDI value from the supervised random forest regression model, building a model response surface.

@@

The same approach was used for the DAI using also a comparison between active bleed size and black staining size as well as missing crown and crown transparency.
The upper boundary for the ranges of active bleed size and black stain size was taken as the centre value between the maximum and minimum and reduced by a factor of eight to give ranges of `r glue("{symptom_ranges$minimum[1]}-{signif(symptom_ranges$center[1] / 8,digits = 3)}mm")` and `r glue("{symptom_ranges$minimum[2]}-{signif(symptom_ranges$center[2] / 8,digits = 3)}mm")` respectively and enable more effective visualisation.
Cases included neutral, moderate and severe with PDI scores greater than 0.5 and DAI scores closest to 0, 0.5 and 1 respectively for AOD and 0, -0.5 and -1 respectively for COD.

@@

A scatter plot of the calculated PDI and DAI scores is shown in Figure # for all 174 trees for which phenotypic descriptor data was collected.
It can clearly be seen that the control trees group towards the lower end of the PDI scale and the symptomatic COD and AOD groups towards the higher end.
This gave a significant difference when PDI of the control and symptomatic (AOD and COD) decline statuses were compared using an independent two sample t-test (`r glue("t({PDI_status_ttest$parameter}) = {signif(PDI_status_ttest$statistic,digits = 3) %>% abs()}, p < 0.001")`) with group means of `r signif(PDI_status_ttest$estimate1,digits = 3)` and `r signif(PDI_status_ttest$estimate2,digits = 3)` for non-symptomatic and symptomatic statuses respectively.

@@

LIME identified crown volume, condition transparency and radius as the top four descriptors contributing to the PDI in each of the healthy, moderate and severe decline cases using lime (Table #).
These descriptors were also the same top descriptors in the overall model importance lists for the PDI.
Increasing decline severity was associated with a reduction in the size and condition of the crown.
For instance, declining trees were found to have an estimated crown volume below 841m^3^.
Healthy trees were found to have a crown condition greater than 71.3%, moderately declining trees between 47.6% and 71.3% and severely declining trees a condition below 23.9%.

@@

The DAI is a much more complex index with more descriptors contributing to it than the PDI such as bleed prevalence, diameter at breast height and crown condition descriptors. 
Simulated response surfaces for the DAI are shown in Figure c. 
These response surfaces show circumstances under which bleed prevalence related descriptors (Figures #a, b & c) and crown condition descriptors (Figures #d, e & f) change for non-symptomatic AOD or COD trees (Figures #a & d respectively), moderate AOD or COD (Figures #b & e respectively) and severe AOD or COD (Figures #c & f respectively). 
Increasing active bleed, staining size and frequency contributed to greater positive values of the DAI and more severe AOD. 
Also higher DAI values could be obtained from trees with greater stature (Figure #c). 
Crown transparency and missing crown have no influence on the DAI for trees of neutral decline type (Figure #d).
Similarly to the PDI, trees with smaller stature and poorer crown condition give lower DAI values (Figures #e & f) and therefore more severe COD.
The relationship of relative tree stature and the acuteness of the decline likely reflects the period of time under which these trees have been afflict. 
AOD occurs over a much shorter period compared COD and therefore has less time in which to inhibit growth of the tree as a whole.

@@

### Suitability of the decline indexes for describing decline severity and type

The phenotypic descriptors were selected for use in this study purposefully

- Suitability of collected phenotypic descriptors and decline indexes for describing decline status
- Index thresholds for category definitions (healthy, declining, severe decline etc.)
- Validation of decline scores - chicken and paradox
- Direct validation difficult 
- Refined definition of remission - repeat measurement

This plot highlights the inaccuracies in manual field classifications of decline status with several trees classified as healthy controls that, when compared quantitatively to other trees, are in fact
in a decline state.

These two indexes provide a two dimensional continuum under which the spectrum of decline severity and type can be assessed.

The two dimensional space provided by these phenotypic indexes, for measuring decline severity is shown in Figure #.

@@

A machine learning approach was used to derive the phenotypic index measures as it allows the moulded to the complex interactions and trends between phenotypic descriptors from which the decline indexes are calculated. 
This would likely be oversimplified using conventional modelling strategies [@breiman2001b].

- reasons for selecting machine learning approach
- use of unsupervised over supervised analysis
- Semi-supervised can be used if unsupervised model is unable to pull out decline severity scale
- For given context, descriptors not contributing to the indexes could be removed
- Concession of model interpretability to predictive capability by using a machine learning approach

Machine learning approaches are usually seen as black box 
- seen as black box but LIME and response surfaces enable interpretation and therefore trust
- LIME can be applied to out of place cases
- Important descriptors identified model-wide aren't necessarily relevant to localised cases.

### Applications and wider scalability

@@

The phenotypic descriptors used to calculate these indexes are easily measurable in the field and require no specialist equipment. 
They allow sensitive comparisons to be made for the same tree between years as well as comparisons of trees between different locations.
which could have great utility in long term forest health monitoring for both forestry management and research.
